The markings of the trade: exploring the patterns of olive oil production in Roman Baetica

### 2020-07-01 Paper/jaccardbaetica.tex
add new changes xavi

The intensification of long-range trade was one of the most important traits of the economy developed duriong the Roman Empire. The development of an extensive road network increased the connectivity between inland communities, but most shipping continued being based on maritime routes particularly in the Mediterranean basin~\citep{temin_market_2001,bevan_mediterranean_2014}. The intensity of this trade between distant regions can be observed both in archaeological and written evidence~\citep{rodriguez_baetican_1998}.

%Material culture is one of the most frequent indicators of trade in the archaeological record. In archaeology, they allow us to highlight a part of the mechanism of production and distribution of goods along the Mediterranean \citep{bevan_mediterranean_2014}. Particularly, the spread of these factors had an important impact during the Roman Age, when the progressive exploitation of communication networks allowed a major interaction between communities \citep{orengo_seeds_2016}\xavi{aquí faltan citas de obras más generales; esto es un ejemplo pero muy específico no?}. 
%spread of goods and ideas had a enormous impact during the Roman Empire. The progressive exploitation of communication networks allowed a major interaction between communities
%This can be seen by the fact that material culture is found in different regions along the Mediterranean as a frequent indicator of trade exchange in the archaeological record.

The Empire developed a series of structures to support and organize this long-range trade and specialized entire regions to massively produce specific goods. An example of this process was the province of \textit{Baetica} (currently Andalusia, southern Spain) which became an important olive oil production centre during\xavi{período?}. Olive oil was an essential good for Romans because it was used in almost every aspect of their daily life such as cooking, hygiene or lighting~\citep{mattingly_d.j._oil_1988}. This high demand required a huge increase on the production which was distributed through amphorae shipped via maritime routes to all the provinces and particularly to Italy and the hundreds of garrisons that the Roman army deployed along its borders~\citep{blazquez_exportacion_1980}.

The structure and processes associated to this massive olive oil production has been extensively discussed over the last decates\citep{rodriguez_economioleicola_1977, Chic_hispania_1997,millet_anforas_1998}. Our knowledge of this economic activity has benefitted from new findings and data sources, but despite all these advances several questions remain open: \xavi{no entiendo esta pregunta} what patterns were followed for the distribution of olive oil to the different provinces? \xavi{no es esta pregunta idéntica a la anterior?} Did each province follow a different pattern for its distribution? \xavi{esta frase no tiene sentido aquí; o la pones antes o en otro párrafo, pero es raro acabar un set de 2 preguntas con una frase que no sea otra pregunta} Nor the lack of written records has made possible to detect any indication of patterns in the olive oil market.

@@

\begin{figure}[htp]
	\centering
\includegraphics[width=\linewidth]{figs/general_map}
\caption{Overview of the sites analyzed in this paper. The color defines the three different regionsu under study (Baetica as producer; Germania and Britannia as consumers) while the size of each dot highlights the number of Dressel-20 amphoric stamps found on each of these sitese}

@@

The landscape of the \textit{Baetican} province was one of the best regions to face the increase in demand of olive oil across the Roman provinces. For this reason the area saw an increase in productivity as a massive infrastructure of olive oil production was gradually deplyoyed. The production and distribution of olive oil grew exponentially\xavi{seguro que exponencial?} for almost three centuries\xavi{qué centuries?} \citep{remesal_concierto}. Hundreds of presses and amphora-making workshops were build near large extensions covered by olive trees. The workshops where amphorae were made and filled with olive oil were places along the rivers Guadalquivir and Genil riverbanks (see Fig.\ref{workshop}).

@@

%\xavi{Estás segura de esto? Porque aquí hay 2 cosas: 1. datar bien y 2 . período de uso. Yo no tengo claro que la mayoría de sellos se usen durante mucho tiempo. En qué está basada la idea? Deberías detallar un poco más este temita cronológico y explicar como sabemos el período de uso.}

@@

The creation of new provinces allowed the Roman Empire the arrival of resources through Mediterranean and Atlantic routes\xavi{no entiendo esto...allowed o required?}. This led to a gradual change in the economic and social structureas the trade networks that supplied the Roman army became more complex. Augustus's administration created the figure of the \textit{praefectura annonae} to organize wheat supply. The role of the \textit{praefectura annonae} was mainly focused on providing through \textit{frumentationes} a fixed monthly amount of wheat to each Roman citizen~\citep{remesal_annona_1986,remesal_concierto}\xavi{pero estabas hablando del ejército no? Has hecho un salto aquí a los ciudadanos}.

@@

%\xavi{este título de sección debería cambiarse por algo más descriptivo. Case studies o algo asínnn? Por otra parte, igual podrías dividirlo en 2 subsections}

%\xavi{en general, echo a faltar un poco más de detalle en esta discusión y en los datos que tienes; en qué se parecen y diferencian las 2 provincias? Si hablas del uso por las campañas o las guarniciones, donde están? Qué strategic points de redistribución hay? Tampoco a saco, pero entender mejor cómo llega el aceite, cómo se redistribuye y quién y dónde se consume}

>>>>

The intensification of long-range trade was one of the most important traits of the economy developed during the Roman Empire. The development of an extensive road network increased the connectivity between inland communities, but most shipping continued being based on maritime routes particularly in the Mediterranean basin~\citep{temin_market_2001,bevan_mediterranean_2014}. The intensity of this trade between distant regions can be observed both in archaeological and written evidence~\citep{rodriguez_baetican_1998}.

%Material culture is one of the most frequent indicators of trade in the archaeological record. In archaeology, they allow us to highlight a part of the mechanism of production and distribution of goods along the Mediterranean \citep{bevan_mediterranean_2014}. Particularly, the spread of these factors had an important impact during the Roman Age, when the progressive exploitation of communication networks allowed a major interaction between communities \citep{orengo_seeds_2016} 
%spread of goods and ideas had a enormous impact during the Roman Empire. The progressive exploitation of communication networks allowed a major interaction between communities
%This can be seen by the fact that material culture is found in different regions along the Mediterranean as a frequent indicator of trade exchange in the archaeological record.

The Empire developed a series of structures to support and organize this long-range trade and specialized entire regions to massively produce specific goods. An example of this process was the province of \textit{Baetica} (currently Andalusia, southern Spain) which became an important olive oil production centre during the Roman Empire. Olive oil was an essential good for Romans because it was used in almost every aspect of their daily life such as cooking, hygiene or lighting~\citep{mattingly_d.j._oil_1988}. This high demand required a huge increase on the production which was distributed through amphorae shipped via maritime routes to all the provinces and particularly to Italy and the hundreds of garrisons that the Roman army deployed along its borders~\citep{blazquez_exportacion_1980}.

The structure and processes associated to this massive olive oil production has been extensively discussed over the last decades\citep{rodriguez_economioleicola_1977, Chic_hispania_1997,millet_anforas_1998}. Our knowledge of this economic activity has benefited from new findings and data sources, but despite all these advances several questions remain open: \xavi{no entiendo esta pregunta} what patterns were followed for the distribution of olive oil to the different provinces? \xavi{no es esta pregunta idéntica a la anterior?} Did each province follow a different pattern for its distribution? \xavi{esta frase no tiene sentido aquí; o la pones antes o en otro párrafo, pero es raro acabar un set de 2 preguntas con una frase que no sea otra pregunta} Nor the lack of written records has made possible to detect any indication of patterns in the olive oil market.

@@

\begin{figure}[htp]
	\centering
\includegraphics[width=\linewidth]{figs/general_map}
\caption{Overview of the sites analyzed in this paper. The color defines the three different regions under study (Baetica as producer; Germania and Britannia as consumers) while the size of each dot highlights the number of Dressel-20 amphoric stamps found on each of these sites}

@@

The landscape of the \textit{Baetican} province was one of the best regions to face the increase in demand of olive oil across the Roman provinces. For this reason the area saw an increase in productivity as a massive infrastructure of olive oil production was gradually deployed. The production and distribution of olive oil grew exponentially\xavi{seguro que exponencial?} for almost three centuries\xavi{qué centuries?} \citep{remesal_concierto}. Hundreds of presses and amphora-making workshops were build near large extensions covered by olive trees. The workshops where amphorae were made and filled with olive oil were places along the rivers Guadalquivir and Genil riverbanks (see Fig.\ref{workshop}).

@@

@@

The creation of new provinces allowed the Roman Empire the arrival of resources through Mediterranean and Atlantic routes\xavi{no entiendo esto...allowed o required?}. This led to a gradual change in the economic and social structures the trade networks that supplied the Roman army became more complex. Augustus's administration created the figure of the \textit{praefectura annonae} to organize wheat supply. The role of the \textit{praefectura annonae} was mainly focused on providing through \textit{frumentationes} a fixed monthly amount of wheat to each Roman citizen~\citep{remesal_annona_1986,remesal_concierto}\xavi{pero estabas hablando del ejército no? Has hecho un salto aquí a los ciudadanos}.

@@
